Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-03-12 manuscript/manuscript.qmd
manuscript_0311

---
title: "Flattering Advice: Avoiding Disappointment in Advice-Giving"
shorttitle: "Flattering Advice"
date: yes
author: 
  - name: "Amanda Chen"
    corresponding: true
    orcid: 0000-0002-4428-627X
    email: zchengj@connect.ust.hk
    affiliations:
      - id: id1
        name: "The Hong Kong University of Science and Technology"
        department: "Department of Management"
        city: "Hong Kong"
  - name: "David Hagmann"
    orcid: 0000-0002-2080-997X
    affiliations:
    - id: id1
# author-note:
  # disclosures:
    # gratitude: The authors are grateful to XXX for their helpful comments.
abstract: Good advice improves decision quality but often requires delivering unpleasant truths that may disappoint advisees. Across three pre-registered and incentivized experiments involving real adviser-advisee interactions (N = 3,900), we show that advisers prioritize avoiding disappointment at the expense of accuracy and their own earnings. In Study 1, advisers financially rewarded for accuracy still tailor recommendations to aspirational goals expressed by advisees, resulting in worse advice. When incentivized to be liked, advisers provide even more flattering advice, and advisees reward this by rating these advisers as more likable, despite the advice being less honest and less accurate (Study 2). The desire to avoid disappointment may lead to inequities if there are differences in expectations across social groups. In Study 3, we examine a setting in which men expect to perform better than women. We show that advisers take into account these expectations, leading to systematically different advice for men and women even when their gender is unknown to advisers. Advisers' efforts to avoid disappointment may thus contribute to systematic gender disparities in advice, with implications for downstream decision-making.

keywords: [Gender difference, Advice, Interpresonal relationship]
bibliography: bibliography.bib
floatsintext: true
mask: false
link-citations: false
format:
  apaquarto-html: default
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    prefer-html: true
  apaquarto-docx:
    default: true
    prefer-html: true
editor: 
  markdown: 
    wrap: 72
---

@@

study1_time1 <- readRDS(here::here("Data/Study1_time1.Rds"))
study1_ability50 <- read.csv(here::here("Data/Cleaned Data/S1_Ability50_clean.csv"))
study1_time2_giver <- readRDS(here::here("Data/Study1_time2_giver.Rds"))
study1_time2_advice <- readRDS(here::here("Data/Study1_time2_advice.Rds"))
study1_time3 <- readRDS(here::here("Data/Study1_time3.Rds"))

@@

@@

Career trajectories for men and women often differ substantially, even
among those with similar qualifications. One source of such differences
may be the advice that people receive. Prior work has documented that
men receive more aspirational advice and women more risk-averse advice,
which may explain why women are less likely to apply for and ultimately
obtain more rewarding positions [@Kanze2018]. Research on gender
discrimination has focused on the role of unconscious bias
[@Greenwald2009; @Greenwald1995], but interventions that try to reduce
this bias have been largely unsuccessful [@Chang2019; @Paluck2009].
Recent work has proposed cognitive, rather than unconscious, biases that
can lead to gender discrimination through the formation of false beliefs
[@Hagmann2024]. Here, we propose a novel source of differences in advice
by drawing on recent work on belief-based utility, expectation-based
reference points [@Koszegi2006; @Koszegi2009], and the hedonic
consequences of information [@Golman2017].

Advice figures prominently in workplace and personal decision-making,
shaping choices about promotions, career paths, and everyday dilemmas
[@Gino2007; @Harvey1997; @Soll2009]. Its fundamental premise is to
improve outcomes by enabling decision-makers to learn from others'
expertise. However, while honest advice can lead to better decisions, it
can also lead to disappointment when it suggests a desired outcome is
unlikely to materialize. Advisers may be motivated to offer flattering
recommendations that align with advisees' hopes, sparing them the
disappointment of a more sobering forecast.

Advice with the potential for disappointment appears in diverse
organizational settings: supervisors deliver feedback on employee
performance, lawyers convey likely outcomes of litigation to their
clients, and academics offer thoughts on colleagues' manuscripts.
Ideally, honest information would help people identify their strengths
and weaknesses, make informed decisions about whether to pursue legal
cases, and strengthen papers prior to submission to academic journals,
thus improving long-term outcomes. In practice, however, interpersonal
considerations may prevent candor. When someone's expectations are (too)
high, an adviser might hesitate to deliver information that undermines
it---particularly if doing so risks negative interpersonal
repercussions.

While much of the literature on advice assumes that advisers aim to provide
the most helpful advice, thier motivations are often more complex [@Yaniv2012;@Dalal2010; @Rader2017]. 
Advisers are not only concerned with the accuracy and usefulness of their advice 
but also with their relationships with advisees. Interpersonal concerns—such
as maintaining a positive image, avoiding conflict, or preserving social 
harmony—can shape the way advice is delivered [@Van2009; @Schaerer2018; @Chentsova2012].
In particular, advisers may strategically soften their messages or withhold 
unfavorable information to manage how they are perceived or to prevent discomfort
for both themselves and the advisee [@Sniezek2001;@Levine2020]. These interpersonal dynamics 
suggest that advice is not always optimized for the recipient’s best interest but
is instead filtered through social and psychological considerations.

Just as advisers may hesitate to provide unwelcome truths, advisees themselves are often 
resistant to hearing them. People often go to great lengths to avoid unpleasant information,
deriving belief-based utility from maintaining favorable views of
themselves [@Golman2017; @Ho2018; @Loewenstein2018]. Moreover, they
often prefer advisers who withhold bad news from them [@Shalvi2019] and
punish those who communicate bad news [@John2019]. Advisers,
anticipating these preferences and consequences face a dilemma: Should
they provide the accurate but potentially painful truth, or shade their
recommendations to preserve the advisee's mood and the adviser's own
standing? Advisers who fear they will be blamed or disliked for
undermining someone's confidence may reasonably avoid candid feedback,
particularly if they don't incur a cost when the advisee makes a
mistake.

@@

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their feedback to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.[^1]

@@

In Study 1, we nudge advisees into a preference for competing against
either a group of top-performers or a group of low-performers, through
the use of a default option. Advisers who observe the actual performance
of advisees and therefore have the relevant information needed to make a
recommendation nonetheless take into account the initial preference when
recommending which group they should compete against. In Study 2, we ask
advisees to upload photos of themselves and advisers rank them on their
attractiveness. Advisers then recommend which rank the advisee should
bet on, receiving an incentive either for the advisee's accuracy or for
being evaluated favorably. Both groups of advisers recommend a rank that
is more attractive than what they themselves have evaluated the advisee,
and those incentivized for likability further inflate the ranking.
Advisees evaluate advisers more favorably when they recommend betting on
a more attractive rank, including viewing them as more trustworthy.

@@

We first examine whether advisers take into account the expectations of
advisees when providing a recommendation. A group of participants
("Advisees") complete a quiz consisting of ten questions that draw on
ego-relevant domains. They then express a non-binding preference to
compete against a group of high performers or low performers on the same
task, based on their past performance. We nudge participants towards
picking either the low or the high performer group by selecting one of
the two by default. A second group of participants ("Advisers") observe
Advisees' score on the quiz and, in one treatment, also which group they
had selected. We predicted that Advisees who were nudged toward the high
performer group would be more likely to receive advice to compete
against this high group when Advisers observed their non-binding choice.

In a preliminary stage, we recruited
`r scales::comma(dim(study1_ability50)[1])` participants from Prolific
and gave them five minutes to complete a quiz consisting of ten items.
The quiz included word puzzles, identifying emotions from photos, and
selecting the best responses for hypothetical scenarios. These questions
were adapted from surveys that measure problem-solving, emotional
intelligence, and communication skills---which we pre-tested as being
important in modern society and hence where participants might have a
stake in doing well. Participants received a bonus of five cents for
each correctly answered question. We then ranked them based on their
score and label the 20 participants with the lowest (highest) scores as
the Low (High) Performer Group.

Next, we recruited `r scales::comma(dim(study1_time1)[1])` participants
for the role of Advisees. They completed the same 10-item quiz, also
earning five cents for each correct answer. We then asked them to
express a non-binding preference for whether they would like to compete
against the High or Low Performer Group. We informed them that they
would be invited back at a later date when they could make a binding
decision and they could earn a bonus based on whether their score on
this quiz was equal to or higher than a randomly selected member from
their comparison group. If they picked the High Performer Group and had
an equal or higher score than a randomly selected member of that group,
they would earn a bonus of 50 cents. If they picked the Low Performer
Group and outperformed a randomly selected member of that group, they
could earn a bonus of 20 cents. If their score was lower in this
comparison, they would not earn an additional bonus. We randomized which
of the two groups (High or Low Performers) was selected by default, and
Advisees were free to select the other group. The survey concluded 
with basic demographic questions.

We then recruited `r scales::comma(dim(study1_time2_giver)[1])`
participants for the role of Advisers, and the focal part of our
experiment. They were informed of the ability quiz that the Advisees had
completed, how the Low Performer Group and High Performer Groups were
constructed, as well as the choice and incentives for the Advisees. We
then asked Advisers to give advice to ten participants on which group
they should compete against. Advisers were randomly assigned to one of
two treatments. In the "Performance" treatment, they observed only the
score of the other participant. In the "Performance + Expectation"
treatment, they additionally observed the advisee's non-binding choice
of which group to compete against. However, since the outcome was based
solely on the past quiz score and advisers, but not advisees, know the
score, this initial choice was not informative for the purpose of
recommending competing against one group or the other. Advisers could
earn the same bonus as one of the participants they had given advice to
and who returned to make a decision. The survey concluded with basic
demographic questions.

Finally, we invited Advisees back for the follow-up survey. Following
our preregistration, we kept the survey open for 7 days. In total,
`r scales::comma(dim(study1_time3)[1])` Advisees returned. The short
survey reminded them of the task they completed in the previous survey,
informed them that other participants from Prolific had observed their
real score and given them advice on which group to compete against, and, 
finally, reminded them of their initial choice of comparison group. 
They were not informed of their true score or the
score of the groups they could compete against. Participants then
observed the advice from a randomly selected adviser and made their
decision.

```{r}
contingency_table <- table(study1_time1$Guess, study1_time1$default)

We begin by examining the performance of the Stage 1 participants.
Because our default treatment takes place after participants completed
the ability quiz, we do not expect a difference in performance across
the Low and High Default treatments. Indeed, the two groups score no
different from one another
(`r papaja::printnum(round(mean(study1_time1[study1_time1$default == "Low",]$Score),2))`
and
`r round(mean(study1_time1[study1_time1$default == "High",]$Score),2)`
for the Low Default and High Default treatments, respectively,
`r papaja::apa_print(t.test(Score ~ default, data = study1_time1, var.equal = T))$statistic`).
As expected, the default manipulation changes their initial choices. In
the "Low Default" treatment, only
`r round(mean(study1_time1[study1_time1$default == "Low",]$Guess),2)*100`%
of participants prefer to compete against the High Performer Group,
compared to
`r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100`%
of participants in the "High Default" treatment ($\chi^2$(1, n = 201) =
36.62, p \< .001).

```{r}
#| label: tbl-study1regs
#| tbl-cap: "Advice to compete against the High Performer Group in Study 1. Displaying the non-binding choice of those for whom the Low Performer Group was selected by default makes it less likely that they are advised to compete against the High Performer Group (Column 1). Column 2 restricts the analysis to advisers who observe advisees' initial non-binding choice and controls for advisees' score on the quiz. Standard errors are clustered at the adviser level."
#| after-caption-space: 0pt

study1_advice <- study1_time2_advice %>%
  dplyr::mutate(Treatment = factor(Treatment, levels = c('Low', 'High'))) %>%
  dplyr::mutate(ExpShow = ifelse(Condition == 'E', 1, 0))

modelsummary::modelsummary(
  list(`All Advice` = lm(Advice ~ ExpShow * Treatment, data = study1_advice), 
       `Expectations Shown` = lm(Advice ~ Treatment + Score, data = study1_advice %>% dplyr::filter(ExpShow == 1))),
  vcov = ~GiverID,
  coef_map = c(
    "TreatmentHigh" = "High Default",
    "ExpShow" = "Expectation Shown",
    "ExpShow:TreatmentHigh" = "Expectation x High Default",
    "Score" = "Score",
    "(Intercept)" = "Constant"
  ),
  stars = TRUE,
  gof_map = list(list('raw' = 'nobs', 'clean' = 'N', 'fmt' = 0))
) %>%
  tinytable::group_tt(
   j = list("Advice to Compete Against High Performers" = 2:3))

We now examine whether advisers take the advisees' expectations into
account. Advisers observe the true performance of the advisee, such that
expectations about performance add no additional information. However,
because participants who score higher are also more likely to pick the
High Performer Group, we use our Default treatment as an instrument that
reflect higher expectations. We show linear probability models for the
decision of the Advisers, which is equal to 1 if they recommend
competing against the High Performer Group and 0 if they recommend the
Low Performer Group. In Column 1 of @tbl-study1regs, we show our main
specificiation with the experimental assignment for Advisees (High
Performer Default vs. Low Performer Default) and Advisers (Expectation
Shown vs. Expectation Hidden), as well as their interaction. Because
each adviser made ten recommendations, we cluster standard errors at the
adviser level. We find that displaying expectations of those in the Low
Performer Default reduces recommendations to compete against the Top
group by 4 percentage points, or about 10% relative to when expectations
are hidden. This decrease is marginally smaller for advisees with a
default choice to compete against the High Performer Group (nudged to
have high expectations, though not necessarily holding them), and whose
advice did not differ based on whether their expectations were shown or
hidden. We reported a linear probability model on advising to compete
against the High Performer group for advisers in the "Expectations"
treatment, controlling for the true performance of the advisee in Column
2 of @tbl-study1regs. We observe a significant main effect of the
default treatment, where a high default increased the likelihood of
receiving flattering advice. It is worth noting that, since only
`r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100`
% of advisees in the high default group actually stuck with their
default choice, this represents a conservative test of the effect of
high expectations on the likelihood of receiving flattering advice.

```{r}
#| label: fig-study1Advice
#| fig-cap: "Advice given in Study 1. Advisees who were nudged to express high expectations were more likely to receive advice to compete against the High Performer Group."

study1_time2_advice %>%
  mutate(Treatment=factor(Treatment, levels=c('High','Low'), labels = c('High default', 'Low default'))) %>%
  group_by(Condition, Treatment) %>%
  summarize(avg_count = mean(Advice), ci = qnorm(0.975)*sd(Advice)/sqrt(dplyr::n())) %>%
  ggplot2::ggplot(ggplot2::aes(x = Condition, y = avg_count, fill = Treatment)) +
  ggplot2::geom_col(position = 'dodge') +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = avg_count - ci, ymax = avg_count + ci), 
                         position = ggplot2::position_dodge(0.9), linewidth = 0.2, width = 0.1) +
  # ggtitle("Advice: Meeting Expectations vs. Absent Expectations") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  scale_y_continuous("P(Compete Against High Performer)",
                     labels = scales::percent) +
  geom_text(aes(x = Condition, y = 0.02,
                label = paste0(round(avg_count*100), "%")),
            position = position_dodge(0.9),
            color = "white") +
  papaja::theme_apa() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        axis.text.x = element_text(size = 12, face = "bold",color = "black"),
        axis.title.y = element_text(size = 12, face = "bold"),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", # No legend
        plot.title = element_text(hjust = 0.5, size = 16)) +
  scale_x_discrete(labels = c("Expectation Shown", "Expectation Hidden")) +
   geom_text(aes(label=Treatment, y = -0.02), position = position_dodge(0.90), color = "#464747"  )

When an adviser knows an outcome that an advisee can only estimate,
learning this estimate should not change the advice that is given.
However, in line with our argument that advisers take into account the
belief utility of the advisee and prefer to avoid disappointing them, we
find that providing information about expectations makes a difference.
Specifically, advisees who expressed a preference for stronger
competitors were more likely to receive flattering advice, encouraging
them to pursue that option. This occurs even though advisers are
incentivized to give good advice.

We next examine if advisers believe that flattering advice will improve
how they are perceived, and if so, whether this perception is true. We
do this by manipulating the incentives for advisers, who receive a bonus
either based on how they are evaluated by an advisee, or a bonus for the
advisee's accuracy.

@@

study2_time1 <- readRDS(here::here("Data/Study2_time1.Rds"))
study2_time1_200 <- readRDS(here::here("Data/Study2_time1_200.Rds"))
study2_time2 <- readRDS(here::here("Data/Study2_time2.Rds"))
study2_finalRank <- readRDS(here::here("Data/Study2_finalRank.Rds"))
study2_time3 <- readRDS(here::here("Data/Study2_time3.Rds"))

@@

We recruit 300 participants from Prolific and, after asking demographic
questions, invite them to upload photos of themselves (selfies) to be
rated by other participants on attractiveness. We obtain selfies from
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women adhering to our instructions (e.g., did not include other people).
In line with our preregistration, we select the first 100 photos from
women to arrive at a gender-balanced sample ($M_{\text{Age}}$ =
`r mean(study2_time1_200$age)` years). Participants are informed that
their selfies will be randomly grouped with those of nine other
participants of their sex and ranked in terms of attractiveness by a
group of new Prolific participants of the opposite sex. They then make
an unincentivized guess of their rank.

Next, we recruit `r scales::comma(dim(study2_time2)[1])` participants
from Prolific for the role of advisers ($M_{\text{Age}}$ =
`r round(mean(study2_time2$age),2)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female). We
first collect demographic information, then match them to a group of the
opposite sex. They rank selfies of ten participants from most to least 
attractive by entering their rankings into a text box next to each picture. 
Because of a limitation with the survey software, we could not validate that each
rank is given only once, and we remove 115 participants who failed to
follow instructions and did not provide a complete ranking.[^2]

@@

TargetsUnique <- study2_time2 %>%
  dplyr::select(Target) %>%
  unique()

TargetsUnique <- merge(TargetsUnique,study2_finalRank, by.x = 'Target', by.y = 'QID7_Id')

@@

```{r}
#| label: fig-study2advice-old
#| fig-cap: "Advisers were asked to suggest what rank to bet on for someone they ranked as the 7th most attractive out of ten people. When incentivized for accuracy, most of them advised betting on 7, which was based on their evaluation of the recipient's attractiveness. However, when incentivized for likability, many participants gave flattering advice by suggesting betting on higher ranks that imply a higher evaluation of attractiveness."
#| eval: false

@@

study2_time2 <- study2_time2 %>% dplyr::mutate(Advice = as.numeric(Advice))

>>>>

---
title: "Flattering Advice: Avoiding Disappointment in Advice-Giving"
shorttitle: "Flattering Advice"
date: yes
author: 
  - name: "Amanda Chen"
    corresponding: true
    orcid: 0000-0002-4428-627X
    email: zchengj@connect.ust.hk
    affiliations:
      - id: id1
        name: "The Hong Kong University of Science and Technology"
        department: "Department of Management"
        city: "Hong Kong"
  - name: "David Hagmann"
    orcid: 0000-0002-2080-997X
    affiliations:
    - id: id1
# author-note:
  # disclosures:
    # gratitude: The authors are grateful to XXX for their helpful comments.
abstract: Advice that improves decision-making does not always feel good. While advice has the potential to improve decision-making by helping advisees more accurately assess feasibility and predict potential outcomes, it can also lead to disappointment when it challenges the advisee’s overly optimistic expectations. Across three pre-registered and incentivized experiments involving real adviser-advisee interactions (N = 3,900), we show that advisers prioritize avoiding disappointment at the expense of accuracy and their own earnings. In study 1, advisers use flattering advice to create more positive impressions, and advisees reward this by rating these advisers as more likable, despite the advice being less honest and less accurate. Even advisers incentivised for accuracy still tailor recommendations to aspirational goals expressed by advisees, resulting in financial loss for them and the advisees (Study 2). The concern for interpersonal consequences may lead to inequities if there are differences in expectations across social groups. In Study 3, we examine a setting in which men expect to perform better than women. We show that advisers take into account these expectations, leading to systematically different advice for men and women even when their gender is unknown to advisers. Advisers' efforts to avoid disappointment may thus contribute to systematic gender disparities in advice, with implications for downstream decision-making.

keywords: [Advice, Interpresonal relationship, Gender difference]
bibliography: bibliography.bib
floatsintext: true
mask: false
link-citations: false
format:
  apaquarto-html: default
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    prefer-html: true
  apaquarto-docx:
    default: true
    prefer-html: true
editor: 
  markdown: 
    wrap: 72
---

@@

@@

<!--# Amanda: I replaced all 'advisers' with 'adviser' for consistency. -->

@@

Advice figures prominently in workplace and personal decision-making,
shaping choices about promotions, career paths, and everyday dilemmas
[@Gino2007; @Harvey1997; @Soll2009]. While others can provide
information in the form of opinions [@Trouche2018], feedback
[@Kluger1996], or personal observations [@Yoon2021] to influence the
decision-maker, this project focuses specifically on advice—defined as
"information that is perceived as intended to help an advisee deal with
their issue of deliberation" [@Milyavsky2024]. Advice is distinct from
other forms of input because its primary purpose is to improve
decision-making by leveraging others' expertise.

Advice has the power to enhance decisions, yet it also carries the risk
of disappointment. For example, a career consultant might advise an
employee to enroll in a management program to strengthen their skill set
before competing for a leadership role. Ideally, such honest advice
helps the employee identify areas for personal growth. In practice,
however, the employee might feel overlooked and frustrated that their
ambitions are not immediately supported. As a result, advisers may be
inclined to offer flattering recommendations that align with—or even
exceed—advisees' hopes, such as encouraging the employee to pursue the
leadership role despite concerns about their readiness. In this way,
interpersonal dynamics may discourage full candor in giving advice.

The objective of this research is to examine the dilemma advisers face:
Should they provide accurate but potentially painful advice, or soften
their recommendations to preserve the advisee's mood and protect their
own standing? Given that advisees tend to prefer advisers who withhold
bad news [@Shalvi2019] nd penalize those who deliver it [@John2019],
advisers may reasonably avoid full candor—especially when they bear no
cost for the advisee’s mistakes. Specifically, when someone's
expectations are (too) high, an adviser may hesitate to provide advice
that challenges them, particularly if doing so risks straining the
relationship or causing negative interpersonal consequences.

Our research contributes to the growing literature on advisers' motives
and intentions in advice-giving. This line of work examines how
advisers' perceived dishonesty or helping intentions influence advice
utilization [@Haran2020; @Mackinger2017; @Milyavsky2024; @Van2009]. By
examining advice within its interactional and communicative context, we
extend this research by incorporating another dimension of advisers'
motives—interpersonal concerns. We demonstrate that advisers adjust
their recommendations to avoid hurting the advisee’s feelings and
self-image. Second, we contribute to the belief-based utility
literature, which suggests that individuals derive satisfaction from
their beliefs about the world and themselves [@Loewenstein2018]. We show
that advisers also place value on preserving advisees' self-perceptions,
often tailoring their advice to align with the advisee’s expectations.
Furthermore, we reveal that this advice-giving strategy has unintended
consequences—it transforms the confidence gap [where women tend to be
less confident than men; @Lundeberg1994] into an advice gap, in which
women are less likely than men to receive advice to aim higher, thereby
reinforcing existing gender inequalities.

## The Motives Behind Advice-Giving

While much of the literature on advice assumes that advisers aim to
provide the most helpful advice, their motivations are often more
complex [@Yaniv2012; @Dalal2010; @Rader2017]. Other than the accuracy
and usefulness of the advice, interpersonal concerns—such as maintaining
a positive image, avoiding conflict, or preserving social harmony—can
shape what advice is delivered and how [@Sniezek2001; @Levine2020]. For
example, research has shown how people often avoid providing negative
feedback directly, opting instead for ambiguity or indirect phrasing to
minimize discomfort [@Levine2020]. These interpersonal dynamics suggest
that advice is not always optimized for the recipient’s best interest
but is instead filtered through social and psychological considerations.

In particular, advisers may strategically soften their messages or
withhold unfavorable evaluations to manage how they are perceived or to
prevent discomfort for both themselves and the advisee [@Van2009;
@Schaerer2018; @Chentsova2012]. When advisers anticipate disagreement or
resistance, they may soften their recommendations or frame them more
diplomatically, sometimes at the expense of accuracy. This suggests that
advice is not merely a means of sharing information but also a tool for
managing social interactions and reducing potential friction [@Van2009].
These interpersonal considerations become even more pronounced when
there is a power imbalance between the adviser and advisee, as
lower-power advisers may be more hesitant to offer direct or critical
guidance to avoid negative repercussions [@Schaerer2018].

Just as advisers may hesitate to provide unwelcome truths, advisees
themselves are often resistant to hearing them. People often go to great
lengths to avoid unpleasant information, deriving belief-based utility
from maintaining favorable views of themselves [@Golman2017; @Ho2018;
@Loewenstein2018]. Moreover, they often prefer advisers who withhold bad
news from them [@Shalvi2019] and punish those who communicate bad news
[@John2019]. Advisers who fear they will be blamed or disliked for
undermining someone's confidence may reasonably avoid candid advice,
particularly if they don't incur a cost when the advisee makes a
mistake.

Providing flattering advice, however, should not be seen as maliciously
misleading the advisee. Rather than being irrational, it tends to be
more optimistic, and this optimism is not necessarily misplaced. Often,
the advisee possesses insights about themselves—such as their motivation
and past experiences—that the adviser may overlook or have no access to.
Take, for example, an employee deciding whether to compete for a
leadership role. While the adviser might focus on the employee’s lack of
management knowledge, the employee may have strong leadership
characteristics, excellent communication skills, and persuasive
abilities. In this way, flattering advice can encourage the advisee to
set ambitious goals, fostering a self-fulfilling prophecy where
confidence drives success [@Jussim1986; @Locke2006].

Of course, if the advice is overly inflated, it may become unhelpful.
However, this is unlikely to happen, as advisers typically aim to boost
the advisee’s confidence while maintaining credibility. Suggesting an
unrealistic option—such as encouraging an entry-level employee to
compete for a CEO position—would come across as insincere or even
mocking rather than flattering.

Beyond protecting the advisee's beliefs and fostering their confidence
and motivation, advisers may also offer flattering advice to shield
themselves from blame if the ultimate decision leads to a poor outcome.
Although both the adviser and advisee recognize the inherent uncertainty
between the advice, the decision, and the final outcome, people tend to
attribute poor outcomes to the adviser [@Weiner1985; @Palmeira2015].
However, research suggests that people differentiate between errors made
intentionally—such as those driven by insufficient knowledge or
deliberate bias—and those made unintentionally, with harsher penalties
imposed on advisers perceived to have deliberately undermined the
decision [@Haran2020]. Flattering advice, even when ultimately
unhelpful, may reduce the likelihood of blame because advisees are less
likely to suspect ill intent from an adviser who appears supportive.
Advisees may instead interpret the poor outcome as an unintended error
rather than deliberate misguidance, particularly because they view the
adviser positively and believe the adviser genuinely holds them in high
regard [@Palmeira2015; @Haran2020]. Even when advisees recognize that
flattering advice was intended to boost their confidence, research shows
that people tend to appreciate such misplaced good intentions, as they
strengthen social bonds by making the gift-giver feel socially closer
[@Zhang2012].

Advisers may choose to inflate their advice for different reasons,
depending on their relationship with the advisee and the stakes of the
decision. For instance, advisers might offer flattering advice to
strangers out of politeness, provide inflated encouragement to friends
to offer emotional support [@Dalal2010], or use their recommendations to
motivate individuals with particularly ambitious or competitive
personalities. We recognize that multiple motives may simultaneously
drive this behavior, and their relative influence—as well as advisers'
overall tendency to inflate advice—may vary across situations. However,
in this project, we focus specifically on whether advisers intentionally
inflate their advice and explore the potential consequences of doing so.

One downstream consequence of giving flattering advice is the potential
to reinforce gender differences — specifically, men are more likely to
receive such inflated advice. Prior work has documented that men receive
more aspirational advice and women more risk-averse advice, which may
explain why women are less likely to apply for and ultimately obtain
more rewarding positions [@Kanze2018]. Research on gender discrimination
has focused on the role of unconscious bias [@Greenwald2009;
@Greenwald1995], but interventions that try to reduce this bias have
been largely unsuccessful [@Chang2019; @Paluck2009]. Recent work has
proposed cognitive, rather than unconscious, biases that can lead to
gender discrimination through the formation of false beliefs
[@Hagmann2024]. Here, we propose that one overlooked mechanism driving
gender disparities in career outcomes may lie in the differential
provision of flattering advice, shaped by the advisee's
self-expectations.

Specifically, when advisers aim to make their advice appear flattering,
they may rely less on an objective evaluation of the advisee's
competence and instead anchor their advice on the advisee’s
self-expectations. Drawing on the literature on belief-based utility and
expectation-based reference points [@Koszegi2006; @Koszegi2009], we
suggest that advisers may align their advice with how ambitious or
confident the advisee appears, so that they give optimistic but not
entirely unrealistic recommendations. This dynamic may inadvertently
disadvantage women. Research shows that men generally hold higher
self-expectations or are more likely to display confidence, while women
tend to exhibit more modest self-assessments despite comparable
competence [@Kanze2018]. As a result, advisers may unintentionally offer
men more ambitious and aspirational advice, encouraging them to pursue
higher-rewarding opportunities, while offering women more cautious,
conservative advice that aligns with their lower expressed
self-expectations. This differential treatment, though seemingly
well-intentioned, can have profound downstream consequences. By
tailoring advice to match the advisee's perceived self-expectations,
advisers may subtly nudge men toward risk-taking, leadership roles, or
competitive opportunities, while directing women toward safer, less
ambitious paths — ultimately reinforcing existing gender disparities in
career outcomes.

# Overview of studies

@@

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their advice to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.[^1]

@@

Across all three studies, we use a three-stage design. In Stage 1, we
measure the advisee’s performance and their self-expectations. In Stage
2, we show the measured performance to the adviser and ask them to
provide advice on how the advisee can maximize their chances of earning
a bonus. Half of the advisers also receive information about the
advisee’s self-expectations. Importantly, this expectation information
is objectively irrelevant to the task, as the bonus is solely based on
the advisee’s performance, which is already measured. While in real
life, self-expectations might contain useful insights (such as
unobservable skills or motivation), in our setup, self-expectations
serve no informative value. Moreover, flattering advice provided in this
setup cannot be justified as a motivational boost (like a pep talk)
because the performance has already been fixed. Finally, in Stage 3, we
re-invite the advisees from Stage 1, show them the advice they received,
and observe how they make their final decisions.

In Study 1, we ask advisees to upload photos of themselves and advisers
rank them on their attractiveness. Advisers then recommend which rank
the advisee should bet on, receiving an incentive either for the
advisee's accuracy or for being evaluated favorably. Both groups of
advisers recommend a rank that is more attractive than what they
themselves have evaluated the advisee, and those incentivized for
likability further inflate the ranking. Advisees evaluate advisers more
favorably when they recommend betting on a more attractive rank,
including viewing them as more trustworthy.

In Study 2, we nudge advisees into a preference for competing against
either a group of top-performers or a group of low-performers, through
the use of a default option. Advisers who observe the actual performance
of advisees and therefore have the relevant information needed to make a
recommendation nonetheless take into account the initial preference when
recommending which group they should compete against.

@@

We first examine if advisers believe that flattering advice will improve
how they are perceived, and if so, whether this perception is true. We
do this by manipulating the incentives for advisers, who receive a bonus
either based on how they are evaluated by an advisee, or a bonus for the
advisee's accuracy.

@@

```{r}

study1_time1 <- readRDS(here::here("Data/Study1_time1.Rds"))
study1_time1_200 <- readRDS(here::here("Data/Study1_time1_200.Rds"))
study1_time2 <- readRDS(here::here("Data/Study1_time2.Rds"))
study1_finalRank <- readRDS(here::here("Data/Study1_finalRank.Rds"))
study1_time3 <- readRDS(here::here("Data/Study1_time3.Rds"))

@@

We recruit 300 participants from Prolific and, after asking demographic
questions, invite them to upload photos of themselves (selfies) to be
rated by other participants on attractiveness. We obtain selfies from
`r scales::comma(dim(study1_time1[study1_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study1_time1[study1_time1$gender == 'Female',])[1])`
women adhering to our instructions (e.g., did not include other people).
In line with our preregistration, we select the first 100 photos from
women to arrive at a gender-balanced sample ($M_{\text{Age}}$ =
`r mean(study1_time1_200$age)` years). Participants are informed that
their selfies will be randomly grouped with those of nine other
participants of their sex and ranked in terms of attractiveness by a
group of new Prolific participants of the opposite sex. They then make
an unincentivized guess of their rank.

Next, we recruit `r scales::comma(dim(study1_time2)[1])` participants
from Prolific for the role of advisers ($M_{\text{Age}}$ =
`r round(mean(study1_time2$age),2)` years;
`r round(mean(study1_time2$gender == "Female")*100, 2)`% Female). We
first collect demographic information, then match them to a group of the
opposite sex. They rank selfies of ten participants from most to least
attractive by entering their rankings into a text box next to each
picture. Because of a limitation with the survey software, we could not
validate that each rank is given only once, and we remove 115
participants who failed to follow instructions and did not provide a
complete ranking.[^2]

@@

TargetsUnique <- study1_time2 %>%
  dplyr::select(Target) %>%
  unique()

TargetsUnique <- merge(TargetsUnique,study1_finalRank, by.x = 'Target', by.y = 'QID7_Id')

@@

```{r}
#| label: fig-study1advice-old
#| fig-cap: "Advisers were asked to suggest what rank to bet on for someone they ranked as the 7th most attractive out of ten people. When incentivized for accuracy, most of them advised betting on 7, which was based on their evaluation of the recipient's attractiveness. However, when incentivized for likability, many participants gave flattering advice by suggesting betting on higher ranks that imply a higher evaluation of attractiveness."
#| eval: false

@@

study1_time2 <- study1_time2 %>% dplyr::mutate(Advice = as.numeric(Advice))
